dispRity ecology demo

### 2017-08-10 doc/dispRity-ecology-demo.Rmd
Minor edits to reflect changes in main manual

This demo aims to give quick overview of the `dispRity` package (v.0.3) for ecological analysis.
Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more details.

To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, we might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.

@@

For this example with ecological data we are going to use data from D. McClean (unpubl.).
This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.

@@

## As well as two list of different factors affecting each experimental plot: treatment and depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
```

@@

A classical way to represent this ordinated data will be to use two dimensional plots.

@@

The first step in such analysis will be to create different subsamples that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
Each of these subsamples contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
In our example, we are going to group the elements according to their depth and treatment.

@@

Second, let’s split the data according to these factors to create the subsamples of the ordinated space by using the `custom.subsamples` function:

@@

Note that the output of `dispRity` functions are `dispRity` objects.
They allow specific attributes to be displayed using the basic S3 functions (`print`, `summary`, `plot`, etc...).
For more details on the `dispRity` objects, see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).

@@

Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity as defined by the user.
For more details on measuring disparity, see the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).

In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.

@@

Similarly than with `custom.subsamples`, `dispRity` displays a `dispRity` object.
To actually look at the calculated values, we can summarise the data by simply using `summary`:

@@

We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
Also, as we can see, each subsamples has different numbers of elements.
It might be interesting to rarefy the data as well to have only subsamples with the same number of elements.
Both steps are easily doable through the `boot.matrix` function.

@@

## Bootstrapping the subsamples and using rarefaction (i.e. only re-sampling 17 elements each time)
(rarefied_data <- boot.matrix(customised_subsamples, bootstraps = 100, rarefaction = 17))
```

We can now rerun a more robust disparity analysis using the bootstrapped data:

@@

Summarising the data
--------------------

We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
For that we will use the `summary` function mentioned above using several of the options specific to `dispRity` objects (see more details with `?summary.dispRity`).

@@

We can also plot the results and have a look at the effect of the number of experimental plots using the generic `plot` function and it's associated arguments (see more details with `?plot.dispRity` for the arguments specific to `dispRity` objects).

@@

As we can see, there seems to be no strong effect of the number of experimental plots in each subsamples (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!

@@

Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.

@@

As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plots and the centroid of the ordinated space.
In other words, it seems that with the second treatment and the second level of depth, the invertebrate communities and composition were further apart (i.e. more dispersed in the ordinated space) than with the first treatment and the first level of depth.

>>>>

This demo aims to give a quick overview of the `dispRity` package (v.0.3) for multidimensional ecological analyses.
Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more detail.

To keep it short, this package allows user to include **all** the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a subset of dimensions.
<!-- For example, we might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats. -->

@@

For this example with ecological data we are going to use data from D. McClean (unpubl.).
The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->

@@

## Create lists of the different factors affecting each experimental plot: treatment and depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
```

@@

A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates. 
<!-- NC: or similar just to make it clearer -->

@@

The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
In our example, we are going to group the elements according to their depth and treatment.

@@

Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:

@@

Note that the output of `dispRity` functions are `dispRity` objects.
They allow specific attributes to be displayed using the basic S3 functions (`print`, `summary`, `plot`, etc...).
For more details on `dispRity` objects, see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).

@@

Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
For more details on measuring disparity, see the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).

<!-- NC: In the manual you use hypervolume - not sure if it matters? -->
In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.

@@

Similarly to the `custom.subsamples` function, `dispRity` displays a `dispRity` object.
To actually look at the calculated values, we can summarise the data by simply using `summary`:

@@

We can also bootstrap the data to test the robustness of the measured disparity to outliers.
Also, as we can see, each subsample has different numbers of elements.
It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
Both steps are easily doable through the `boot.matrix` function.

@@

## Bootstrapping the subsamples and using rarefaction 
## (i.e. only re-sampling 17 elements each time)
(rarefied_data <- boot.matrix(customised_subsamples, bootstraps = 100, rarefaction = 17))
```

We can now run a more robust disparity analysis using the bootstrapped data:

@@

Summarising the results
-----------------------

We can now summarise the results using various options such as the confidence intervals and the central tendency.
For that we use the `summary` function mentioned above, using several of the options specific to `dispRity` objects (see more details with `?summary.dispRity`).

@@

We can also plot the results and look at the effect of the number of experimental plots using the generic `plot` function and its associated arguments (see more details with `?plot.dispRity` for the arguments specific to `dispRity` objects).

@@

As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!

@@

Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.

@@

As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plot and the centroid of the ordinated space.
In other words, it seems that with the second treatment and the second level of depth, the invertebrate communities and composition were further apart (i.e. more dispersed in the ordinated space) than with the first treatment and the first level of depth.

<!-- NC: Some comment here about how this is different to the result you get with 2D methods alone? -->
